---
title: "PCOS analysis"
format:
  revealjs:
     slide-number: true
     embed-resources: true
     theme: moon
      
author: "Agnes Lorenzen, Cecille Hobbs, Freja E. Klippmann, Julie Dalgaard Petersen & Mille Rask Sander"
---

```{r}
# | echo: true 
# | eval: true  

library(tidyverse)
```

## Introduction {.smaller}

### Background

-   Polycystic ovary syndrome (PCOS) is a syndrome documented in women in their menustruating ages

-   Documented symptoms are often; period pains, irregular periods, ovary related problems and hormone imbalance

-   Patients with PCOS often have problems with pregnancy and potential complication with/in pregnancy

-   However, it is still not verified what the cause of PCOS is.

### Aim

The aim of this study is to examine a data set (found on Kaggle) of patients with and without PCOS. The data set has been made in India and data comes from 10 different hospitals.

## Data handling approach {style=".smaller"}

::: columns
::: {.column width="50%"}
::: {style="font-size: 0.7em"}
-   <strong>Raw data</strong>:<br />541 observation divided into 45 variables.<br />
-   <strong>01_load_data</strong>:<br />Simply loads the data.<br />
-   <strong>02_clean_data</strong>:<br />
    -   Fixing random cells and replacing them with NA
    -   Rename & factorizing columns
    -   Split dataframe into body and blood measurements
    -   Removed empty column.<br />
:::
:::

::: {.column width="50%"}
::: {style="font-size: 0.7em"}
-   <strong>03_augment</strong>:<br />
    -   Unit changes ( inch to cm)

    -   Rounding & grouping BMI

    -   Change Blood type and cycles from numeric values to characters

    -   Create new column for cycle/ pregnancy stage

    -   Merging data frame into one file<br />

        </div>

        <div style="font-size: 0.6em">

``` r
# Rounding of BMI and dividing into categories
body_measurements <- body_measurements |>
  mutate(BMI = round(BMI, 1)) |> 
  mutate(BMI_class = case_when(
    BMI < 18.5 ~ "Underweight",
    BMI <= 18.5 | BMI < 25 ~ "Normal weight",
    BMI <= 25 | BMI < 30 ~ "Overweight",
    BMI >= 30 ~ "Obesity")) |>
  mutate(BMI_class = factor(BMI_class,
                            levels =  c("Underweight", 
                                        "Normal weight",
                                        "Overweight", 
                                        "Obesity"))) |>
  relocate(BMI_class, .after = BMI)
```
:::
:::
:::

## Data handling approach {style=".smaller"}

+---------------------------------------------------------+----------------------------------------------------------------------+
| 02 Clean data                                           | 03 Augment data                                                      |
+=========================================================+======================================================================+
| -   ~Fixing random cells and replacing them with NA~    | -   ~Unit changes ( inch to cm)~                                     |
| -   ~Rename & factorizing columns~                      | -   ~Rounding and\ grouping BMI~                                     |
| -   ~Split data frame into body and blood measurements~ | -   ~Change Blood type and cycles from numeric values to characters~ |
| -   ~Removed empty column~                              | -   ~Create new column for cycle/ pregnancy stage~                   |
|                                                         | -   ~Merging data frame into one file~                               |
+---------------------------------------------------------+----------------------------------------------------------------------+

## Descriptive analysis of data

::: columns
### Left Column

```{r}
# | echo: true 
# | eval: true  

PCOS <- read_tsv(file = "../data/PCOS_merged.tsv")

PCOS_dim <- PCOS |>
  dim() |>
  tibble()|>
  rename("PCOS dimensions" = "dim(PCOS)")|>
  print()

PCOS |> 
  count(PCOS_diagnosis) |>
  as.tibble() |>
  print()

```

::: {.column width="50%"}
### Right Column
:::
:::

## Analysis 1

::: {.column width="50%"}
hh

hh
:::

## Analysis 2

her

## PCA of blood measurements

her

## PCA of body measurements

her

## Discussion

her

## Conclusion

-   no significance
